Contracts for unions and intersections are hard for Nickel

### 2021-05-31 paper.tex
Add conclusion

>>>>

The design and implementation of programming languages is a challenging task. It
is a delicate exercise to balance complexity, expressiveness, efficiency and
user experience all together. Unfortunately for designers, a language cannot
usually be reduced to a simple sum of small independent features. They are
rather composed of a lot of moving parts that interact in unexpected ways. The
more a language grows, the more any new addition has to be weighted carefully,
as it has the potential to interact with any of the existing features.

Part of recent developments revolve around mixing static typing and dynamic
typing via gradual typing, in order to get the best of both worlds. Contracts
(or casts in the gradual typing terminology) are crucial to this endeavour by
allowing for a safe interaction between typed and untyped code. But we also made
a case that contracts are interesting in themselves in a fully dynamically typed
language (and even in a fully statically typed language).  Contracts are a
precious idiom for data validation or to check pre- and post-condition that
would be hard to ensure statically. In spirit, contracts are similar to
assertions, but they are part of a principled framework, they are composable,
they have good error reporting and they are integrated as a language feature. In
particular, contracts natively handles higher-order functions and built-in data
structures. This makes them an great candidate for integration in the Nickel
configuration language.

This is only the beginning of the story. A language designer then has to choose
what contract combinators to provide among a number of possibilities: there is
the standard package including higher-order function contracts, record contracts
and list contracts.  Then, one could add polymorphic
contacts~\cite{BlameForAll}, dependent function
contracts~\ref{DependentContracts}, and so on. For validation purpose, union and
intersection contracts are really appealing. Unions have a intuitive
interpretation as modeling alternatives and thus appear implicitly or explicitly
in a lot of data structures.

Alas, the implementation of union and intersection contracts is not as simple as
it first appears. In conjunction with higher-order contracts, or even just basic
contracts in a lazy language, they behave in a way that is reminiscent of
side-effects. These behaviors were, for most of them, already known: we do not
claim their discovery. However, as language implementers, we have made an actual
attempt at including them in a concrete language. We exposed what, in our
opinion, are the most salient flaws and their consequences. Even in a pure
language, union and intersection contracts break the equivalence between
call-by-name and call-by-value, and consequently can prevent CSE-like
optimizations. There are discrepancies between the expected and actual semantics
that make some behaviors unintuitive. Their implementation is a non-trivial
change that contaminates the evaluation of all other contracts, and the
performance penalty is not obvious to assess.

To conclude, let us make clear that we do not think union and intersection
contracts are fundamentally broken, that they can not be implemented correctly,
or that they do not bear any value (quite the contrary). They may still make
sense to have in a language, and some apparent difficulties in the
implementation could be lifted some day. But as often, there are gaps between
the theoretical framework, a proof-of-concept, a prototype, and the integration
in an actual language. We hope that our attempt may serve as a cautionary tale,
that for union and intersection contracts, these gaps are sometimes large in
places.
